---
title: "hw8"
format: 
  html:
    code-fold: true
---



## Exercise 1. Supply and demand

```{r}
library(vegawidget)
```




```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.json"
  },
  "columns": 2,
  "concat": [
    {
      "title": "1",
      "width": 200, "height": 120,
      "transform": [
        {"fold": ["capacity", "demand"], "as": ["key", "value"]}
      ],
      "mark": "bar",
      "encoding": {
        "x": {"field": "date", "type": "temporal", "timeUnit": "yearmonth"},
        "y": {"field": "value", "type": "quantitative"},
        "xOffset": {"field": "key", "sort": "descending"},
        "color": {"field": "key"}
      }
    },
    {
      "title": "2",
      "width": 200, "height": 120,
      "layer": [
        {"mark": "line", "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "capacity", "type": "quantitative"}, "color": {"datum": "capacity"}}},
        {"mark": "line", "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "demand", "type": "quantitative"}, "color": {"datum": "demand"}}}
      ]
    },
    {
      "title": "3",
      "width": 200, "height": 120,
      "layer": [
        {"mark": {"type": "bar", "size": 8}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "demand", "type": "quantitative"}, "color": {"datum": "demand"}}},
        {"mark": {"type": "bar", "size": 13}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "capacity", "type": "quantitative"}, "color": {"datum": "capacity"}}}
      ]
    },
    {
      "title": "4",
      "width": 200, "height": 120,
      "layer": [
        {"mark": {"type": "bar"}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "demand", "type": "quantitative"}, "color": {"datum": "demand"}}},
        {"mark": {"type": "bar"}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "capacity", "type": "quantitative"}, "color": {"datum": "capacity"}}}
      ]
    },
    {
      "title": "5",
      "width": 200, "height": 120,
      "layer": [
        {"mark": {"type":"rule", "strokeWidth": 3, "stroke": "grey"}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "capacity", "type": "quantitative"}, "y2": {"field": "demand"}}},
        {"mark": {"type": "point", "fill": "white", "size": 40}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "capacity", "type": "quantitative"}, "color": {"datum": "capacity"}}},
        {"mark": {"type": "point", "fill": "white", "size": 40}, "encoding": {"x": {"field": "date", "type": "temporal"}, "y": {"field": "demand", "type": "quantitative"}, "color": {"datum": "demand"}}}
      ]
    },
    {
      "title": "6",
      "width": 200, "height": 120,
      "transform": [
        {"calculate": "max(0, datum.demand - datum.capacity)", "as": "unmet"}
      ],
      "layer": [
        {
          "mark": "line",
          "encoding": {
            "x": {"field": "date", "type": "temporal"},
            "y": {"field": "unmet", "type": "quantitative"}, "color": {"datum": "capacity"}
          }
        },
        {
          "mark": {"type": "point", "filled": true},
          "encoding": {
            "x": {"field": "date", "type": "temporal"},
            "y": {"field": "unmet", "type": "quantitative"}, "color": {"datum": "capacity"}
          }
        }
      ]      
    }
  ],
  "config": {
    "axis": {"title": null},
    "range": {
      "category": ["mediumblue", "deepskyblue"] 
    }
  }
}' |> as_vegaspec()
```


## Exercise 2. Jobs

**example graphic**
```{r}
'{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "spec": {
    "mark": {"type": "point"},
    "transform": [
      {"pivot": "year", "value": "perc", "groupby": ["job", "sex"]}
    ],
    "encoding": {
      "x": {
        "field": "1850",
        "type": "quantitative",
        "scale": {"constant": 0.00001, "type": "symlog"}
      },
      "y": {
        "field": "2000",
        "type": "quantitative",
        "scale": {"constant": 0.00001, "type": "symlog"}
      },
      "tooltip": [
        {"field": "job"},
        {"field": "sex"},
        {"field": "1850"},
        {"field": "2000"}
      ]
    }
  },
  "facet": {"column": {"field": "sex", "title": "", "type": "nominal"}},
  "title": {
    "text": "Percent of people working various jobs",
    "subtitle": "Each dot represents the one occupation. Hover to see which occupation it is."
  },
  "data": {
    "url": "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
  }
}' |> as_vegaspec()
```

#### a. Create a graphic that shows the sum of all the values of perc in each year separated by men and women. Explain what your graphic reveals about what perc means.

```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
  },
  "width": 400,
  "height": 200,
  "mark": "line",
  "encoding": {
    "x": {"field": "year", "type": "quantitative"},
    "y": {
      "aggregate": "sum",
      "field": "perc", "type": "quantitative"
    },
    "color": {"field": "sex"}
  }
}' |> as_vegaspec()
```
`perc` seems to be representing the percentage of each job and sex relative to the total population in that year.

#### b.

```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "transform": [
    {"filter": {"field": "year", "oneOf": [1850, 1900, 1950, 2000]}},
    {"pivot": "sex", "value": "perc", "groupby": ["job", "year"]}
  ],
  "facet": {"field": "year", "type": "nominal", "title": "Year"},
  "columns": 2,
  "spec": {
    "width": 200,
    "height": 200,
    "mark": {"type": "point"},
    "encoding": {
      "x": {
        "field": "men",
        "type": "quantitative",
        "scale": {"constant": 0.00001, "type": "symlog"},
        "axis": {"format": ".0%"},
        "title": "Percentage of Men"
      },
      "y": {
        "field": "women",
        "type": "quantitative",
        "scale": {"constant": 0.00001, "type": "symlog"},
        "axis": {"format": ".0%"},
        "title": "Percentage of Women"
      },
      "tooltip": [
        {"field": "job"},
        {"field": "men"},
        {"field": "women"}
      ]
    }
  },
  "title": {
    "text": "Percent of men and women working various jobs",
    "subtitle": "Comparing employment share by gender across years. Each dot represents the one occupation. Hover to see which occupation it is."
  },
  "data": {
    "url": "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
  }
}' |> as_vegaspec()
```

#### c.

```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "transform": [
    {"pivot": "sex", "value": "perc", "groupby": ["job", "year"]},
    {"filter": {"field": "year", "oneOf": [1850, 1900, 1950, 2000]}},
    {
      "joinaggregate": [
        {"op": "sum", "field": "men", "as": "total_men_perc"},
        {"op": "sum", "field": "women", "as": "total_women_perc"}
      ],
      "groupby": ["year"]
    },
    {"calculate": "datum.men / datum.total_men_perc", "as": "prop_men"},
    {"calculate": "datum.women / datum.total_women_perc", "as": "prop_women"}
  ],
  "facet": {"field": "year", "type": "nominal", "title": "Year"},
  "columns": 2,
  "spec": {
    "width": 200,
    "height": 200,
    "mark": {"type": "point"},
    "encoding": {
      "x": {
        "field": "prop_men",
        "type": "quantitative",
        "scale": {"type": "symlog", "constant": 0.0001},
        "axis": {"format": ".0%"},
        "title": "Within-Men Percentage"
      },
      "y": {
        "field": "prop_women",
        "type": "quantitative",
        "scale": {"type": "symlog", "constant": 0.0001},
        "axis": {"format": ".0%"},
        "title": "Within-Women Percentage"
      },
      "tooltip": [
        {"field": "job"},
        {"field": "prop_men", "format": ".01%", "title": "% of All Men"},
        {"field": "prop_women", "format": ".01%", "title": "% of All Women"}
      ]
    }
  },
  "title": {
    "text": "Comparing Career Concentrations Within Genders",
    "subtitle": ["Each dot represents an occupation. Axes show the percentage of men and women employed in that job,","relative to their respective total gender populations."]
  },
  "data": {
    "url": "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
  }
}' |> as_vegaspec()
```




## Exercise 3. Price at the pump

```{r}
'{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://raw.githubusercontent.com/hjyi01/others/refs/heads/main/pump_price_for_gasoline_us_per_liter.csv"
  },
  "transform": [
    {
      "fold": ["1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016"],
      "as": ["year_raw", "price_raw"]
    },
    {"calculate": "toNumber(datum.price_raw)","as": "price"},
    {"calculate": "toNumber(datum.year_raw)","as": "year"},
    {
      "lookup": "country",
      "from": {
        "data": {"url": "https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv"},
        "key": "name",
        "fields": ["region", "sub-region"]
      }
    },
    { "filter": "isValid(datum.price)"},
    {
      "aggregate": [{"op": "mean", "field": "price", "as": "avg_price"}],
      "groupby": ["region", "year"]
    }
  ],
  "width": 400,
  "height": 250,
  "mark": {"type": "line"},
  "encoding": {
    "x": {
      "field": "year", 
      "type": "quantitative", 
      "axis": {"format": "y"},
      "title": "Year"
    },
    "y": {
      "field": "avg_price", 
      "type": "quantitative", 
      "title": "Average Price (USD)"
    },
    "color": {"field": "region", "type": "nominal", "title": "Region"},
    "tooltip": [
      {"field": "year"},
      {"field": "region"},
      {"field": "avg_price", "title": "Average Price (USD)", "format": "0.2f"}
    ]
  },
  "title":{
    "text": "Gasoline Pump Price Trend by Region"
  }
}' |> as_vegaspec()
```

#### Do any countries fail to get matched?

Yes, some of the countries failed to get matched. This is because of the different notations of countries. For example, the same country is named `South Korea` in gas prices (US $ per liter) from GapMinder and `"Korea, Republic of"` in countries, regions, and codes data.